\documentclass[../main.tex]{subfiles}
\graphicspath{{\subfix{../figures/}}}

\begin{document}
\chapter{Acknowledgements}
This entire body of work would have been impossible without the tireless support of past and present members of the Wallace lab.
In a literal sense, this work would not have existed without the experiments conducted with expertise and insight by Jamie and Weronika on top of their own projects during a period of global turmoil.
I will be eternally grateful and I want you to know how exceptional you are.
Although, having never worked in the lab, I have to trust the work was done in yeast. 

I like to thank the students I've mentored over the years for their incredible work and for reminding me how cool science is: Abhishak Jain, Hanqin Du, Jingqi Chong, and Xuejia Ke. 
The long list of lab members who have created a warming and humorous community over the years: Clémence, Savvas, Ania, Louis, Apple, Amy, Domenico, Flic, Liz, and Rachael.
Laura and Rosey who guided me through my crash course in molecular biology from the first day I entered the office.
Marah for always finding time to listen and for providing ample opportunities to consume alcohol.

I would also like to thank my co-supervisor Guido Sanguinetti for patiently answering my questions during the early days and for helping to push it over the line at the end.
And finally, my supervisor Edward Wallace for gambling on a stellar astrophysicist as his first PhD student in an experimental fungal transcriptomics lab. 
I hope it paid off. 

An honourable mention has to go to my parents and sister -- and their pets that I am highly allergic to -- who have had to endure me during this time.
Your whisky and treats are in the post. 
I will leave the allocation to Hector.
As for all the friends and flatmates -- and half of the baristas in the Lothian area -- who have had the pleasure of my company over the last four years, you're welcome.
\end{document}